---
title: "RNG of Weibulls"
author: "Greg Johnson"
output: pdf_document
---


Consider the Weibull distribution:

$$f_\Theta(\theta)=\frac{\alpha}{\beta^\alpha}\theta^{\alpha-1}\exp \big( -(\theta/\beta)^\alpha \big)$$

How can we generate realizations from this distribution "by hand"?

Let $\alpha=2$ and $\beta=0.5.$ One method is to sample from Weibull using the *inverse CDF method*.

$$F_\Theta(\theta) = \int_0^\theta f_\Theta(\theta)d\theta = \frac{\alpha}{\beta^\alpha} \int_0^\theta \theta^{\alpha-1} \exp \big( -(\theta/\beta)^\alpha \big)$$
Formally, we would use u-substitution but it's fairly obvious what the integral is:

$$F_\Theta(\theta) =\Bigg[- \exp \big( -(\theta/\beta)^\alpha \big) \Bigg]_0^\theta = 1 - \exp \big( -(\theta/\beta)^\alpha \big)$$
Now we invert the CDF:

$$p=1 - \exp \big( -(\theta/\beta)^\alpha \big)$$
$$\log (1-p) = -(\theta/\beta)^\alpha$$
$$F^{-1}_\Theta(p)=\beta \big[-\log (1-p) \big]^{1/\alpha}$$
Now we apply the Inverse CDF method to sample from the Weibull with $\alpha=2$ and $\beta=0.5.$ For b samples, repeat this b times:

1. $U^* \sim Unif[0,1]$
2. $X^*=F^{-1}_\Theta(U^*)$

```{r tidy=TRUE}
#set Weibull parameters
alpha = 2 
beta = 0.5

invWeibull = function(p){beta*(-log(1-p))^(1/alpha)}
invWeibull = Vectorize(invWeibull)

Weibull_samples = invWeibull(runif(1000))
  
hist(Weibull_samples,col="cornflowerblue",xlab="Reals",
     main="Histogram of Weibulls",freq=FALSE,breaks=30)
curve(dweibull(x,2,.5),add=TRUE,col="red",lwd=2,lty=2)
```

Now let's use *rejection sampling* to sample from the Weibull distribution with $\alpha=2$ and $\beta=0.5.$ Specifically, use the Exponential distribution with scale $\beta$ and multiplier $k=\alpha$ as the envelope.

First, let's show graphically that this envelope adequately covers the Weibull distribution for all $\theta.$
```{r tidy=TRUE}
curve(dweibull(x,2,.5),col="cornflowerblue",from=0,to=1.1,lwd=2,xlab="X",ylab="Density",main="Rejection Sampling Densities")
curve(2.05*dexp(x,2),add=TRUE,lwd=2,col="red",lty=2)
legend("topright",legend=c("Weibull","Envelope"),lwd=2,lty=c(1,2),col=c("cornflowerblue","red"))
```

Great! Now we can proceed with the Rejection Sampling algorithm:

1. Sample $\theta^* \sim Exp(\beta)$
2. Sample $u^* \sim U[0,1]$
3. If $u^*<\frac{f(\theta^*)}{G(\theta^*)}$ then keep $\theta^*.$ Otherwise reject it.

```{r tidy=TRUE}
alpha = 2
beta = .5
B = 1000

weibull_samples_RS = numeric()

for(b in 1:B){
  theta = rexp(1,2)
  u = runif(1)
  if(u<dweibull(theta,2,.5)/(2.01*dexp(theta,2))){
    weibull_samples_RS = c(weibull_samples_RS,theta)
  }
}

hist(weibull_samples_RS,col="cornflowerblue",xlab="Reals",
     main="Histogram of Weibulls",freq=FALSE,breaks=30)
curve(dweibull(x,2,.5),add=TRUE,col="red",lwd=2,lty=2)
```

